Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-10-01 analysis/paper/paper.Rmd
test PAT
revise introdution and change the title

---
title: "Exploring pottery specialization during the colonial contact in the 17-19th century in northeast Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Pottery production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as whole when specialized production was organized that hints changes in production systems. Craft specialization can provide further evidence about the type of production organization, and the presence of elite control (Arnold 2000; Blackman, et al. 1993; Costin 1991; Hirshman, et al. 2010). The underlying assumption is that the specialized mass production of ceramics will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers (Arnold 2000: 334).

We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (AD 600-100) presenting a high consistency in form throughout the region compared to other places in the island.

Here we examine pottery measurements and shapes from Kiwulan, a large multi-component archaeological site in NE Taiwan, to investigate if there are any changes resulting from different periods of foreign contact, European and then Chinese, that might indicate social changes in the indigenous society. We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels? Did the pottery production become more standardized and homogeneous in shape after foreign contact? Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network? This is important because the

# Background and Research area

@@

In this study, I examine 291 pots recovered from Kiwulan site.  The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

@@

# Acknowledgments

@@

>>>>

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies. We explore the culture changes in indigenous societies in northeast Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production. Pottery production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as whole when specialized production was organized that hints changes in production systems. Based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers, we may be able to determine the type of production, and the presence of elite control, from artefact analysis (Arnold 2000: 334; Blackman, et al. 1993; Costin 1991; Hirshman, et al. 2010).

We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization. We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels? Did the pottery production become more standardized and homogeneous in shape after foreign contact? Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network?

Pottery specialization can be identified by examining the standardization of pottery measurements (citation). However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization. Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan. We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.

We find differences in shape and shape standardization of pottery that indicate changes in pottery production resulting from foreign contact, suggesting increasing craft specialization and changes in local social organization at Kiwulan. These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes. In addition, our case study, which includes an openly available research compendium of R code suitable for use with any other assemblage, will help to expand the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.

# pottery shape analysis 
- review about pottery shape analysis recently

# Research area
Kiwulan is located in the north part of Yilan plain in northeast Taiwan. Yilan is

@@

In this study, we examine 291 pots recovered from Kiwulan site.  The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

@@

# Acknowledgements

@@

Word count: `r wordcountaddin::word_count()`
